---
title: "Density Plots"
author: "Amanda White"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 1
    number_sections: true
  pkgdown:
    toc: true
    toc_depth: 1

vignette: >
  %\VignetteIndexEntry{Density Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, 
  fig.height = 6
)
```

Density plots display the distribution of meta-data values corresponding to peaks observed in one or more samples, a group, or pair of groups.  The `densityPlot` function can display histograms or smoothed density, and it can display both together if plotting a single distribution (i.e. one sample or one aggregated group). The distribution of any numeric column of the `e_meta` element may be plotted.

To construct a density plot for a single sample, first construct a subset containing only the sample of interest. By default, `densityPlot` plots the distribution (without a histogram) of the chosen variable.

```{r message=FALSE}
library(ftmsRanalysis)
data("exampleProcessedPeakData")

one_sample <- subset(exampleProcessedPeakData, samples="EM0011_sample")
densityPlot(one_sample, variable="NOSC")
```

Setting the `plot_hist` parameter to `TRUE` will display both the curve and histogram.

```{r}
densityPlot(one_sample, variable="NOSC", plot_hist=TRUE)
```

Alternatively, using `plot_curve=FALSE` gives just a histogram. The `yaxis` parameter controls the scale of the y-axis: it can be either "density" (default) or "count"

```{r}
densityPlot(one_sample, variable="NOSC", plot_hist=TRUE, plot_curve=FALSE, yaxis="count")
```

The `densityPlot` function can also choose a single sample from a larger dataset, using the `samples` parameter:

```{r}
densityPlot(exampleProcessedPeakData, variable="NOSC", samples="EM0069_sample")
```

The `samples` parameter may also be used to specify more than one sample, although only curves may be plotted, not  histograms.

```{r}
densityPlot(exampleProcessedPeakData, variable="NOSC", samples=c("EM0011_sample", "EM0013_sample", "EM0015_sample"),
            curve_colors=c("blue", "red", "green"))
```

Since the `exampleProcessedPeakData` object has a group definition (see `group_designation` for how to construct this), the `groups` parameter can be used to select multiple groups to plot (these curves are the distribution combining all peaks observed by samples in the group).

```{r}
getGroupDF(exampleProcessedPeakData)
densityPlot(exampleProcessedPeakData, variable="NOSC", samples=FALSE, groups=c("M_C", "M_S"))
```

The `samples` and `groups` parameters work in similar ways to select multiple distributions to plot. An `NA` value indicates that all samples or groups in the data object should be plotted, a `FALSE` value suppresses curves, and a character vector may be used to select specific samples and/or groups. The `samples` parameter is `NA` by default (so it will plot all samples found) and `groups` is `FALSE` by default (do not plot groups). 

The easiest way to plot all samples in a group, and the group aggregate is to subset down by group, then call `densityPlot`. (Recall that all samples are plotted by default so only the `groups` parameter must be specified.)

```{r}
MSgroup <- subset(exampleProcessedPeakData, groups="M_S")
densityPlot(MSgroup, variable="NOSC", groups=NA)
```
