---
title: "Kendrick Plots"
author: "Amanda White"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 1
    number_sections: true
  pkgdown:
    toc: true
    toc_depth: 1

vignette: >
  %\VignetteIndexEntry{Kendrick Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, 
  fig.height = 6
)
```

A Kendrick plot displays the Kendrick defect versus Kendrick mass for each peak observed. Kendrick plots provide a visualization which allows one to sort peaks in a spectra (sample) by their homologous relatives. Compounds that differ from one another by a known structural unit (e.g. CH<sub>2</sub>) can be identified in the visualization by peaks plotted on horizontal or diagonal straight line.

The `ftmsRanalysis` package 
provides a method to construct these plots, with a variety of options which are described below. 

# Single Sample Kendrick Plots

To construct a Kendrick plot for a single sample, first construct a subset containing only the sample of interest. By default the `kendrickPlot` function will color points according to their class using the `bs1` definition. (Please see the vignette on [Van Krevelen plots]("van_krevelen_plots.html") for a description of the class calculation options and differences.)

```{r message=FALSE}
library(ftmsRanalysis)
data("exampleProcessedPeakData")

one_sample <- subset(exampleProcessedPeakData, samples="EM0011_sample")
kendrickPlot(one_sample)
```

The resulting plot is a [`plot_ly`](https://cran.r-project.org/package=plotly) object and is interactive. Click on a name in the legend on the right to toggle whether those points are visible. Double-click any name to show only the corresponding points. Mouse over a point to see its molecular formula and mass. Other buttons for zooming, panning and exporting to PDF become visible at the top when the mouse is over the plot area.

There are parameters to customize many of the visible options of this plot. For example to customize the colors, the `colorPal` parameter accepts a color palette function such as `col_factor` in the [`scales`](https://cran.r-project.org/package=scales) package. The `vkBoundarySet` parameter controls which of the 3 boundary sets are used to calculate class; valid options are `bs1` (default), `bs2`, or `bs3`.

```{r}
library(scales)
cat_names <- rownames(getVanKrevelenCategoryBounds("bs1")$VKbounds)
color_palette <- col_factor(palette = "Set1", domain=cat_names)

kendrickPlot(one_sample, colorPal=color_palette, vkBoundarySet = 'bs2')
```

The points may be colored by another column of molecular property by using the `colorCName` parameter. By default the function constructs an appropriate color palette for the chosen data type (integer, numeric or factor) but one may be specified using the `colorPal` parameter. Black points indicate observed peaks for which the chosen meta-data column is NA.

```{r}
kendrickPlot(one_sample, colorCName = "NOSC", title="Van Krevelen Plot Colored by NOSC Value")
```

```{r}
kendrickPlot(one_sample, colorCName="N", title="Van Krevelen Plot Colored by Number of N")
```

The points may also be colored by an `e_data` column such as abundance, transformed abundance or presence/absence. If abundance is used, it will be log-transformed prior to display because the distribution of abundance values typically has a long right tail making the colors difficult to interpret. 

```{r}
kendrickPlot(one_sample, colorCName="EM0011_sample", legendTitle="log(Abundance)")
```


# Group Van Krevelen Plots

Another use of the `kendrickPlot` function is to plot treatment groups. For example, one can construct a group subset, calculate the number or proporation of samples in which each peak was observed (using `summarizeGroups`) and color the points by that summary.

```{r, message=FALSE}
ms_group <- subset(exampleProcessedPeakData, groups="M_S")
ms_summary <- summarizeGroups(ms_group, summary_functions=c("n_present", "prop_present"))
ms_summary$f_data
```

```{r}
kendrickPlot(ms_summary, colorCName="M_S_prop_present", title="M_S Group Colored by Proportion of Times Observed", 
             legendTitle = "Proportion<br>Present")
```

For group plots, the points may still be colored by meta-data columns, and peaks that are observed in any sample in the group will be displayed.

# Group Comparison Van Krevelen Plots

It can be useful to compare peaks that are unique to one treatment group versus another. This can be done by constructing a group comparison object and using `summarizeGroupComparisons` to determine group uniqueness, as shown below.

Clicking the "Observed in Both" label on the right hides those points to focus on the points that are unique to M_S or M_C.

```{r}
m_groups <- divideByGroupComparisons(exampleProcessedPeakData, comparisons=list(c("M_S", "M_C")))[[1]]$value
comp_summary <- summarizeGroupComparisons(m_groups, summary_functions="uniqueness_gtest", 
                                            summary_function_params=list(
                                              uniqueness_gtest=list(pres_fn="prop", pres_thresh=0.3,
                                                                    pvalue_thresh=0.05)
                                          ))
summary(comp_summary$e_data)

kendrickPlot(comp_summary, colorCName="uniqueness_gtest", title="Comparison of M_S and M_C Groups") 
```

