---
title: "Interactivity Between Plots"
author: "Amanda White and Sarah Reehl"
date: "5/14/2018"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This vignette shows how to link two Plotly plots produced by fticRanalysis functions in Shiny such that 
selections in one plot can be highlighted in the other. See http://plotly-book.cpsievert.me/linking-views-with-shiny.html 
for additional information.

# Some important Plotly features

There are a couple of features in Plotly that are not typically used but are crucial for linking plots in Shiny. 

## Source
The first is the concept of a "source". The source feature of a plot is just a character value that is used to identify
the, well, *source* of a mouse event. Setting this is important if you want to distinguish where your mouse event came 
from. There are two ways to do this, the first is in the `plot_ly` function, just specify a source parameter, e.g. 
`myplot <- plot_ly(source="plot1", ...)`. In this example, we're using Plotly objects that are the output
of a function, so we're not calling `plot_ly` directly in this vignette. But we can set the source field after 
constructing the plot like this: `myplot$x$source <- "plot1"`. 

## Key
Another useful concept is that of a key. Each data object (point, bar, line, etc) in a Plotly plot can have its own key. 
The key is specified in the `add_markers` or `add_trace` or similar function. E.g.
```{r background_example, eval=FALSE}
p <- plot_ly(data=plot_data, x=~x, y=~y) %>%
      add_markers(key=~key, color=~color)
```

The keys corresponding to the selected items will be part of the event data structure that is available after a
mouse event occurs. Keys appear to always be translated to character by Plotly, regardless of their original data type.
In the `fticRanalysis` package, the two scatter plot functions `vanKrevelenPlot` and `kendrickPlot` 
use the EDataColname field as the key for each point. (See the `fticRanalysis::getEDataColname` function.) The `densityPlot`
function produces a histogram, where the key for each bar is a 2-element vector of the min and max x-axis values for the bar.

## Plotly mouse events

Plotly generates mouse events for selection, click and hover actions. Mouse event information is stored in Shiny's 
`input` variable with special names that indicate the source and mouse event type. The names take the form 
`.clientValue-`, the mouse event type, then the source value. Mouse event types are `plotly_selected`, `plotly_click` and `plotly_hover`. For example, mouse event names look like: `.clientValue-plotly_selected-source1` and 
`.clientValue-plotly_click-source2`.

Once added to `input` these event data fields are not removed with each mouse event, but only the relevant
fields are updated. This means we need a way (beyond presence or absence of those `input` fields) to keep track of the 
source of the last mouse event. To do this we'll create a reactive value and use the `shiny::observeEvent` function to populate
its value. 

Each mouse event data item is a data frame with a row for each item affected (selected/clicked/hovered) and the following columns:

* `curveNumber`: indicates the data trace number
* `pointNumber`: the point within the specified curve
* `x`: x value
* `y`: y value
* `key`: key provided during plot construction

Plotly provides a convenience method, `event_data` that retrieves this mouse event data frame for a given source and 
event type. This function pulls the data from the relevant slot of `input`.


# Example: two scatter plots

The first example demonstrates the linkage of two scatter plots: a Van Krevelen plot and a Kendrick plot. 
Points selected in either plot will be highlighted in the other. 

First, we generate the two plots from our fticRanalysis example data.

```{r example1a, message=FALSE, warning=FALSE}
library(magrittr)
library(fticRanalysis)
library(plotly)
devtools::load_all("/Users/d3l348/Files/MinT/github/fticRanalysis/")

data(peakIcrProcessed)
.data <- subset(peakIcrProcessed, samples="EM0011_sample")

p1 <- vanKrevelenPlot(.data, title = "Van Krevelen Plot of EM0011", vkBoundarySet = "bs1") %>%
  layout(dragmode="select", showlegend=FALSE)

p2 <- kendrickPlot(.data, title="Kendrick Plot of EM0011") %>%
  layout(dragmode="select", showlegend=FALSE)
```

Notice that in both cases, we've used the layout function to set the drag mode to "select", this tells plotly to use the mouse 
click and drag to select point rather than zoom. (We've also hidden the legends to conserve space.)

Next we're going to give both plots "source" characteristics which will be used to determine from which plot a mouse event originated.

```{r example1b}
p1$x$source <- "vk_source"
p2$x$source <- "kendrick_source"
```

Here we'll create a reactive value to track the source of the last mouse event, and use the `observeEvent` function 
to populate its value. When each type of event is observed, the reactive value `lastEvent` is set accordingly.

```{r example1c}
lastEvent <- reactiveValues(source="none")

# Observe plotly-selected event from vk_source
observeEvent(input$`.clientValue-plotly_selected-vk_source`, {
  lastEvent$source <- "VK"
}, priority = 10)

# Observe plotly-selected event from kendrick_source
observeEvent(input$`.clientValue-plotly_selected-kendrick_source`, {
  lastEvent$source <- "Kendrick"
}, priority = 10)
```

The next step is to define the `output` plot elements and write code to look at the most recent Plotly mouse event to render
selected data in cyan.

Implicit in this code is the knowledge that the Van Krevelen and Kendrick plotting functions set the data key to be the
EDataColname value (the `Mass` column for this dataset). This means that the `key` element of the event data will contain
keys (Masses) associated with the selected points.

```{r example1d}
output$vk_plot1 <- renderPlotly({
  last_event <- lastEvent$source

  if (last_event == "VK") {
    d <- event_data("plotly_selected", source="vk_source")
    if (!is.null(d)) {
      tmp_dat1 <- dplyr::filter(.data$e_meta, Mass %in% d[["key"]])
      p1 <- p1 %>% 
        add_markers(x=~OtoC_ratio, y=~HtoC_ratio, data=tmp_dat1, 
                    marker=list(color="cyan"), name="Selected" )
    }
  } else if (last_event == "Kendrick") {  
    d2 <- event_data("plotly_selected", source="kendrick_source")
    if (!is.null(d2)) {
      tmp_dat1 <- dplyr::filter(.data$e_meta, Mass %in% d2[["key"]])
      p1 <- p1 %>% 
        add_markers(x=~OtoC_ratio, y=~HtoC_ratio, data=tmp_dat1, 
                    marker=list(color="cyan"), name="Selected" )
    }
  }  
  p1
})

output$kendrick_plot1 <- renderPlotly({
  last_event <- lastEvent$source

  if (last_event == "VK") {
    d <- event_data("plotly_selected", source="vk_source")
    if (!is.null(d)) {
      tmp_dat1 <- dplyr::filter(.data$e_meta, Mass %in% d[["key"]])
      p2 <- p2 %>% 
        add_markers(x=~kmass, y=~kdefect, data=tmp_dat1, 
                    marker=list(color="cyan"), name="Selected" )
    }
  } else if (last_event == "Kendrick") {  
    d2 <- event_data("plotly_selected", source="kendrick_source")
    if (!is.null(d2)) {
      tmp_dat1 <- dplyr::filter(.data$e_meta, Mass %in% d2[["key"]])
      p2 <- p2 %>% 
        add_markers(x=~kmass, y=~kdefect, data=tmp_dat1, 
                    marker=list(color="cyan"), name="Selected" )
    }
  }  
  p2
})

```

Finally, add some Shiny output elements to display the plots.

```{r example1e }
fluidRow(
  column(6, plotlyOutput('vk_plot1')),
  column(6, plotlyOutput('kendrick_plot1'))
)
```

# Example: a scatter plot and a histogram

The second example shows highlighting selections between a Van Krevelen plot and a histogram. We'll make another Van Krevelen
plot like Example 1, and when points are selected there, the distribution of their NOSC values will be 
shown on the histogram plot. When a bar of the histogram is selected via a mouse click, the corresponding points will be rendered 
in cyan on the scatter plot.

```{r example2a}

p3 <- vanKrevelenPlot(.data, title = "Van Krevelen Plot of EM0011", vkBoundarySet = "bs1") %>%
  layout(dragmode="select", showlegend=FALSE)
p3$x$source <- "vk_source2"

p4 <- densityPlot(.data, variable = "NOSC", yaxis="count", smooth_curve = FALSE, title="NOSC Distribution for EM0011") 
p4$x$source <- "hist_source"


# Set up reactive flag to keep track of what was the last mouse event
lastEvent2 <- reactiveValues(source="none")
observeEvent(input$`.clientValue-plotly_selected-vk_source2`, {
  lastEvent2$source <- "VK"
}, priority = 10)
observeEvent(input$`.clientValue-plotly_click-hist_source`, {
  lastEvent2$source <- "hist"
}, priority = 10)
```

In the scatter plot functions `vanKrevelenPlot` and `kendrickPlot`, the EDataColname values (the `Mass` column for this dataset)
are used as point keys. In the density plot, the key corresponding to each bar is set to be a two-element vector of that bar's 
min and max. This is used to map from the mouse click to the max and min values on which to filter points for the scatter plot. 

It's also important to note that the output of `densityPlot` has an attribute called `hist_data` which contains the output
of the `hist` command used to generate the barplot. This is important because it contains the locations of each bar, allowing
us to add another set of bars for the selected points at the same widths and locations as the original, rather than estimating
bar sizes from just the selected data.

```{r example2b}
output$vk_plot2 <- renderPlotly({
  last_event <- lastEvent2$source
  
  if (last_event == "VK") {
    d <- event_data("plotly_selected", source="vk_source2")
    if (!is.null(d)) {
      tmp_dat1 <- dplyr::filter(.data$e_meta, Mass %in% d[["key"]])
      p3 <- p3 %>% 
        add_markers(x=~OtoC_ratio, y=~HtoC_ratio, data=tmp_dat1, 
                              marker=list(color="cyan"), name="Selected" )
    }
  } else if (last_event == "hist") {  
    d2 <- event_data("plotly_click", source="hist_source")
    if (!is.null(d2)) {
      bounds <- as.numeric(d2$key[[1]])
      tmp_dat1 <- dplyr::filter(.data$e_meta, as.numeric(NOSC) >= bounds[1] & as.numeric(NOSC) < bounds[2])
      p3 <- p3 %>% 
        add_markers(x=~OtoC_ratio, y=~HtoC_ratio, data=tmp_dat1, 
                              marker=list(color="cyan"), name="Selected" )
    }
  }  
  p3
})

output$hist_plot <- renderPlotly({
  hist_data <- attr(p4, "hist_data")
  last_event <- lastEvent2$source

  if (last_event == "VK") {
    d <- event_data("plotly_selected", source="vk_source2")
    if (!is.null(d)) {
      tmp_dat <- dplyr::filter(.data$e_meta, Mass %in% d[["key"]])
      hist_breaks <- hist_data$x-0.5*hist_data$barwidth
      hist_breaks <- c(hist_breaks, tail(hist_breaks,1)+tail(hist_data$barwidth, 1))

      tmp_hist <- hist(tmp_dat$NOSC, breaks=hist_breaks)
      p4 <- p4 %>% add_bars(x=tmp_hist$mids, y=tmp_hist$counts, width=diff(tmp_hist$breaks), 
                     marker=list(color="cyan")) %>%
        layout(showlegend=FALSE)
    }
  } else if (last_event == "hist") { 
    d2 <- event_data("plotly_click", source="hist_source")
    if (!is.null(d2)) {
      tmp_dat <- hist_data[d2$pointNumber+1, ]
      p4 <- p4 %>% add_bars(x=~x, y=~count, width=~barwidth, data=tmp_dat, 
                     marker=list(color="cyan")) %>%
        layout(showlegend=FALSE)
    }
  }
  p4
})

fluidRow(
  column(6, plotlyOutput('vk_plot2')),
  column(6, plotlyOutput('hist_plot'))
)
```

